---
title: "Association Rules"
author: "Ming-Yu Liu"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: pygments
---

<style type="text/css">
p{ /* Normal  */
   font-size: 18px;
}
body{ /* Normal  */
   font-size: 18px;
}
td {  /* Table  */
   font-size: 14px;
}
h1 { /* Header 1 */
 font-size: 32px;
}
h2 { /* Header 2 */
 font-size: 26px;
}
h3 { /* Header 3 */
 font-size: 22px;
}
code.r{ /* Code block */
  font-size: 14px;
}
pre { /* Code block */
  font-size: 14px
}
</style>



The goal of unsupervised learning is to discover previously unknown patterns (easy) that are also interesting (more challenging) from data. In most cases unsupervised learning isn’t strong enough to provide much value alone, but can provide value for knowledge discovery in highly exploratory settings. Particularly knowledge discovery of data quirks and structure. **Association analysis** is one branch of unsupervised learning that takes the form of "When X occurs, Y also occurs Z% of the time". While generating rules is a straightforward procedure that requires little to no knowledge of the underlying data, post-processing large collection of rules to identify the interesting and valid ones often requires a lot more effort.


# Association Rules with Tabular Data

In market basket analysis problems, association rules can be useful and drive business decisions by themselves. But the same algorithm can be applied to traditional tabular data (spreadsheets or tables from a relational database), where each row is treated as a transaction and all transactions have the same number of items (one per column). Items are represented by the variable name and the value of that variable for the given row. For example, one row from a five column dataset of students could be represented as: {finalMathGrade=B, eyeColor=brown, state=Minnesota, playSports=YES, inBand=NO}. Traditional association rules run on categorical data. If a table contains continuous data of interest, it can be discretized into bins for analysis.

This is a handy first-step approach to develop a deeper understanding of a new dataset that we're not famailiar with and inform more sophisticated analyses. As the lists of association rules is like a lists of facts about the data at hand.

Some scenarios where it can be helpful:

**Learn About The Data**

This is especially important if we're unfamiliar with the dataset that was given to us. In this case, rules with a perfect confidence of 1 will explain ground truths in your data that are usually indicative of business rules, which would be difficult to uncover without documentation or a domain expert. For example when column V1 takes tha value of A or B, column V5 will always equals C. Since association rules handles only categorical variables, imputation or exclusion of missing values to accommodate numerical computation is not necessary. Missing values are commonly coded as just another possible category rather than discarded.   

Although most of the rules will simply expose database quality issues rather than “big picture” issues, it can be a productive first step. After generating the association rules, we can then home in on interesting rules and generate hypotheses to investigate further. Likewise, rules can always be pruned if they are uninteresting.

**Learn The Domain**

Association rules are often (and rightly) criticized for generating gobs of trivial and redundant rules. However, rules that are trivial to a seasoned domain expert might provide new and useful information to a data scientist jumping into a new domain. While often not the final answer, association rules can be used to start the conversation and knowledge sharing process between analysts, domain experts and stakeholders. We often find ourselves in the chicken-egg scenarios where domain experts ask us (the consulting data scientists) what we need to know. But we often know so little about the domain and the data context that we don’t even know how to start asking intelligent questions. Association rules can be that starting point.

# Apriori

We'll use the the Titanic dataset as our toy dataset. This is a 4-dimensional
table with summarized information on the fate of passengers (survival) on the Titanic according to their social class, sex, age.

```{r, message=FALSE, warning=FALSE}

library(DT)
library(arules)
library(data.table)
setwd('/Users/ethen/programming/association')

load('titanic.raw.rdata')
dt <- data.table(titanic.raw)
dt

```

`apriori` takes a transaction class data or any data format that can be transformed to it (e.g. data.frame or matrix), and a parameter argument that takes a list of arguments to specify the parameter for the algorithm.

```{r}

rules <- apriori( 
	dt,

	# the min/max len denotes the min/max number of items in a itemset
	parameter = list( support = 0.1, confidence = 0.5, minlen = 1, maxlen = 5 ),

	# don't print the algorthm's training message
	control = list( verbose = FALSE )
)

# inspect the generated rules, which returns a data.frame
# including the left/right hand side, support, confidence and lift
inspect( head(rules) )

```

For appearance we can specify we only want rules with rhs containing "Survived" only (we then specfiy the default parameter for all the other items that are not expliciting specified).

```{r}

rules <- apriori( 
	dt,
	parameter = list( support = 0.1, confidence = 0.5, minlen = 1, maxlen = 5 ),
	appearance = list( rhs = c( 'Survived=No', 'Survived=Yes' ), default = 'lhs' ),
	control = list( verbose = FALSE )
)

# we can sort the rules
rules_sorted <- arules::sort( rules, by = 'lift' )
inspect(rules_sorted)

```

## Redundant Rules

```{r}

inspect(rules_sorted[1:2])

```

Rule #2 is a redudant rule since it provides no extra knowledge in addition to rule #1, which already tells us that all 2nd-class children survived. 

```{r}

# find redundant rules and prune them
subset_matrix <- arules::is.subset( rules_sorted, rules_sorted )
subset_matrix[ lower.tri( subset_matrix, diag = TRUE ) ] <- NA
redundant <- colSums( subset_matrix, na.rm = TRUE ) >= 1
which(redundant)

rules_pruned <- rules_sorted[!redundant]
inspect(rules_pruned)

```


# Reference

- [Blog: Association rule analysis beyond transaction data](http://brooksandrew.github.io/simpleblog/articles/association-rules-beyond-transactional-data/)
- [RDataMining slides: Association Rule Mining](http://www.rdatamining.com/docs/association-rule-mining-with-r)

http://www.almaden.ibm.com/cs/quest/papers/sigmod93.pdf

http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.53.2780&rep=rep1&type=pdf

https://cran.r-project.org/web/packages/arules/vignettes/arules.pdf

https://cran.r-project.org/web/packages/arulesViz/vignettes/arulesViz.pdf